The Overlapping Control Volume Finite Element Method for Multiphase Porous Media Flow Modelling \\ Part I: Numerical Formulation

### 2015-01-07 Papers/Prototype_1_Method/article.tex
updates

\usepackage[obeyFinal]{todonotes}

@@

\title{The Overlapping Control Volume Finite Element Method for Multiphase Porous Media Flow Modelling \\
Part I: Formulation}

@@

\begin{abstract} 
This is the first of two papers devoted to a novel finite element method formulation to solve subsurface multi-fluid flows. The overlapping control volume finite element method (OCVFEM) has two major strengths, firstly it uses a dual consistent pressure-velocity representation in control volume and finite element spaces which ensures local mass conservation. Secondly, it uses a family of finite element types that ensures Darcy equations for multiphase flows can be exactly enforced. The underlying mass conservation equations are solved in control volume space and finite elements are used to obtain high-order fluxes on control volume boundaries. The novelty of the method lies in (a) permitting both continuous and discontinuous description of pressures and saturations between elements; (b) the use of arbitarily high-order polynomial representation for pressure and velocity within the mixed formulation; (c) the use of high-order flux-limited methods to avoid introducing non-physical oscillations as well as achieving high-order time accuracy where and when possible.   In this paper, the new model is described and test cases are introduced to show the numerical accuracy in one- and two-dimensional simulations. More extensive applications are introduced in the second part of this two part paper. 
\end{abstract}

@@

Finite difference methods (FDM) have been extensively used in modelling and simulation of fluid flows in porous media~\citep{aziz_1986, chen_1997, chen_2005} however, they are strongly dependent on mesh quality and orientation and cannot easily represent complex geometries. In addition, FDM produce excessive numerical dispersion in heterogeneous media \citep{chavent_1986}. The geometrical flexibility of finite element methods (FEM) has proven to overcome these deficiencies. Among FEM-based formulations for porous media, the control volume finite element method (CVFEM) has become increasingly popular as it can guarantee local mass conservation and high-order numerical accuracy as well as being able to use tetrahedral geometry-conforming elements~\citep{forsyth_1990, cordazzo_2004, fung_1992, geiger_2004, cumming_2011, hurtado_2007}.  \citet{huber_2000} demonstrated that the vertex-centred finite volume - finite element method (or Box scheme) can achieve similar goals~\citep[see also][]{helmig_1997}.

For porous media flows, CVFEM discretisations with Galerkin-FEM formulation for the pressure equation has been used to solve Darcy and Richard equations as described by \citet{fung_1992} and \citet{cumming_2011}. \citet{durlofsky_1993,durlofsky_1994} investigated a mixed FEM formulation (i.e., piecewise constant in pressure and piecewise linear in velocity) and classical CVFEM for single phase Darcy flow. He concluded that the latter is more computationally efficient and numerically accurate than the former for simple problems involving a (relatively) small number of degrees of freedom. Moreover, for complex discontinuous and heterogeneous problems, mixed FEM formulations lead to physically realistic solutions.

>>>>

%\usepackage[obeyFinal]{todonotes}
\usepackage{todonotes}

@@

\title{The Overlapping Control Volume Finite Element Method for Multiphase Porous Media Flow Modelling \\
Part I: Numerical Formulation}

@@

\begin{abstract} 
This is the first of two papers devoted to a novel finite element method formulation to solve subsurface multi-fluid flows. The overlapping control volume finite element method (OCVFEM) has two major strengths, firstly it uses a dual consistent pressure-velocity representation in control volume and finite element spaces which ensures local mass conservation. Secondly, it uses a family of finite element types that ensures multiphase Darcy equations can be exactly enforced. The underlying mass conservation equations are solved in control volume space and finite elements are used to obtain high-order fluxes on control volume boundaries. The novelty of the method lies in (a) permitting both continuous and discontinuous description of pressures and saturations between elements; (b) the use of arbitarily high-order polynomial representation for pressure and velocity within the mixed formulation; (c) the use of high-order flux-limited methods to avoid introducing non-physical oscillations as well as achieving high-order time accuracy where and when possible. In this paper, the new model is described and test cases are introduced to show the numerical accuracy in one- and two-dimensional simulations. More extensive applications are introduced in the second part of this two part paper. 
\end{abstract}

@@

Finite difference methods (FDM) have been extensively used in modelling and simulation of fluid flows in porous media~\citep{aziz_1986, chen_1997, chen_2005} however, they are strongly dependent on mesh quality and orientation and cannot easily represent complex geometries. In addition, FDM produce excessive numerical dispersion in heterogeneous porous matrix \citep{chavent_1986}. The geometrical flexibility of finite element methods (FEM) has proven to overcome these deficiencies. Among FEM-based formulations for porous media, the control volume finite element method (CVFEM) has become increasingly popular as it can guarantee local mass conservation and high-order numerical accuracy as well as being able to use tetrahedral geometry-conforming elements~\citep{forsyth_1990, cordazzo_2004, geiger_2004, hurtado_2007}.  \citet{huber_2000} demonstrated that the vertex-centred finite volume - finite element method (or Box scheme) can achieve similar goals~\citep[see also][]{helmig_1997}.

%For porous media flows, CVFEM discretisations with Galerkin-FEM formulation for the pressure equation has been used to solve Darcy and Richard equations as described by \citet{fung_1992} and \citet{cumming_2011}. 
\citet{durlofsky_1993,durlofsky_1994} compared the performance of a mixed FEM formulation (i.e., piecewise constant in pressure and piecewise linear in velocity) and the classical CVFEM for single phase Darcy flow. He concluded that the latter is more computationally efficient and numerically accurate than the former for simple problems involving a (relatively) small number of degrees of freedom. In addition, for complex discontinuous and heterogeneous problems, mixed FEM formulations often lead to physically realistic solutions.

### 2015-01-07 Papers/Prototype_1_Method/article_figure1.tex
updates

  \begin{center}
\vbox{\hbox{\hspace{1cm}
    \includegraphics[width=0.8\textwidth]{compar-dg.eps}}
   \hbox{ \hspace{1cm} \includegraphics[width=0.8\textwidth]{compar-dg-bdt.eps}}}
    \caption{Advection problem test-cases: control volume solution (5
      finite elements with constant width of 0.2 are used) with
      Courant number of 0.005 (top) and 0.5 (bottom).  The solution
      has been advected a distance of 0.2 with 200 (top) and 2
      (bottom) time steps. The 50 elements solution is shown to give
      an indication of the accuracy of the solutions and has a Courant
      number (based on the element width) of 0.005.\label{compar-dg}}
  \end{center}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=1.2\textwidth]{compar-dg-bdt-diff}
\end{center}
\caption{Advection problem test-cases: discontinuous (between
  elements) control volume formulation within 5 finite elements (width
  of 0.2). Courant number is of 0.5 (or 2 if based on the minimum
  width of a control volume). The solution has been advected a
  distance of 0.2 with 2 time steps. The 50 element solution is for
  pure advection and is shown to give an indication of the accuracy of
  the solutions and has a Courant number (based on the element width)
  of 0.005.\label{compar-dg-bdt-diff}}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\vbox{\hbox{\hspace{-1.3cm}
\includegraphics[width=17.5cm,height=12.5cm]{theta-bdt}}
\vspace{0.cm}}
\caption{Advection problem test-cases: value of $\theta$-time stepping
  parameter on the faces of the control volumes at the end of the
  simulation -- 5 FE are used and the Courant number (based on the
  element width of 0.2) is 0.5 (or 2 if based on the minimum width of
  a control volume). The solution has been advected a distance of 0.2
  length units with 2 time steps.}
\label{theta-bdt}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\vbox{\vspace{-2cm}
\hbox{\hspace{-.5cm}
\includegraphics[width=14.0cm,height=10.cm]{converg-cv}}
\vspace{-.5cm}
\hbox{\hspace{-.5cm}
\includegraphics[width=14.0cm,height=10.cm]{converg-fem}}
}\vspace{-.3cm}
\caption{Advection problem test-cases: convergence of the solutions
  with discontinuity between elements and with increased resolution. A
  pure advection problem (from left to right) - the Courant number is
  0.005. The solution has been advected a distance of 0.2. Control
  volume solutions (top) and FEM interpolation (bottom) of the control
  volume solutions are shown. \label{converg}}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\hbox{\hspace{-1cm}
\includegraphics[width=17.5cm,height=12.5cm]{converg-compare-fem}}
\caption{Advection problem test-cases: comparison of convergence of
  the solutions with and without discontinuity between elements and
  with an upwind and central difference flux between the elements and
  with increased resolution (dashed line: 5 elements, full line: 20
  elements). A pure advection problem (from left to right) - the
  Courant number is 0.005. The solution has been advected a distance
  of 0.2. The FEM interpolation of the control volume solutions are
  shown. \label{converg-compare-fem}}
\end{figure}

@@

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=1.\textwidth]{bl-exact-meth-upwind.eps}
\end{center}
\caption{Buckley--Leverett test-cases: Saturation solutions for the
  continuous upwind method for different 1D P$_{1}$DG-P$_{2}$ mesh
  resolutions and comparison against standard analytical solution.
\label{bl-exact-meth-upwind}}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
  %\begin{center}
\vbox{\hbox{\hspace{2.5cm}
    \includegraphics[width=0.62\textwidth]{BL_1d_P0DGP1_convergence.eps}}
\vspace{-.0cm}\hbox{\hspace{2.5cm}
    \includegraphics[width=0.62\textwidth]{BL_1d_P1DGP2_convergence.eps}}
\vspace{-.0cm}\hbox{\hspace{2.5cm}
    \includegraphics[width=0.62\textwidth]{BL_1d_P2DGP3_convergence.eps}}}
   % \includegraphics[width=0.45\textwidth]{BL_2d_P1DGP2_convergence}
    \caption{Buckley--Leverett test-cases: Saturation profiles for a
      number of element pairs and numerical resolutions in 1D --
      P$_{0}$DG-P$_{1}$ (top), P$_{1}$DG-P$_{2}$ and P$_{2}$DG-P$_{3}$
      (bottom).\label{fig:BL_profiles}}
  %\end{center}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\vbox{\hbox{\hspace{1.cm}
    \includegraphics[width=0.8\textwidth]{L1_convergence_rate.eps}}
\vspace{.0cm}\hbox{\hspace{1.cm}
    \includegraphics[width=0.8\textwidth]{L2_convergence_rate.eps}}}
    \caption{Buckley--Leverett test-cases: L1 (top) and L2 (bottom)
      error convergence rates for a number of element
      pairs. \label{fig:BL_converg-rates}}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=1.\textwidth]{bl-upwind-v-up-and-down.eps}
\end{center}
\caption{Buckley--Leverett test-cases: Comparison of the optimal
  upwind formulation when using upwinding (OU) and coupled
  upwind/downwind (OU-D). The finite element interpolation of the
  saturation field $\left(S_{1}\right)$ is shown at different mesh
  resolutions. Downwind seems to detract from the accuracy of the
  solution. \label{bl-upwind-v-up-and-down}}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\vbox{
\begin{center}
\includegraphics[width=1.\textwidth]{bl-exact-meth-cv-0-8-ele50.eps}
\end{center}
\vspace{0.cm}}
\caption{Buckley--Leverett test-cases: Comparison of control volume
  solutions using 80$\%$ upwinding and with optimal upwinding and
  using 50 continuous P$_{1}$DG-P$_{2}$
  elements. \label{bl-exact-meth-cv-0-8-ele50}}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=1.\textwidth]{bl-dg-2eles.eps}
\end{center}
\caption{Buckley--Leverett test-cases: Two element solution using the
  discontinuous formulation. Saturation field from both CV solution
  and FEM interpolation are shown.  \label{bl-dg-2eles}}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\vbox{
\hbox{\hspace{.3cm}\includegraphics[width=.9\textwidth]{bl-dg-4-10-20.eps}}
\vspace{-0.cm}
\hbox{\hspace{.3cm}\includegraphics[width=.9\textwidth]{bl-dg-cent-4-10-20.eps}}}
\caption{Buckley--Leverett test-cases: Saturation field obtained from
  the discontinuous and continuous formulation with different mesh
  resolutions. Solutions with (top) and without (bottom) upwinding
  scheme. Notice that oscillations are suppressed with the upwinding
  scheme.\label{bl-dg-cent-4-10-20}}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\vbox{
\hbox{\hspace{.3cm}\includegraphics[width=.9\textwidth]{bl-dg-4-10-vers-cty.eps}}
\vspace{-0.cm}
\hbox{\hspace{.3cm}\includegraphics[width=.9\textwidth]{bl-dg-p1-2-4-5-10-20-40.eps}}}
\caption{Buckley--Leverett test-cases: Saturation field obtained from
  (top) continuous and discontinuous (between elements) formulations
  (solution with 50 elements may be considered as a converged
  result). Solution obtained (bottom) from linear pressure (P1)
  formulation with different mesh resolution with comparison against
  P2-pressure formulation (continuous). \label{bl-dg-4-10-vers-cty}}
\end{figure}

\begin{figure}[H]
\vbox{
\begin{center}
\includegraphics[width=17.5cm,height=12.5cm]{./doc_figures/bl-dg-4-10-vers-cty}
\end{center}
\vspace{0.cm}}
\caption{Gas saturations shown comparing the accuracy of the
  discontinuous between elements and continuous formulation. The 50
  element continuous solution may be viewed as a converged result.  }
\label{bl-dg-4-10-vers-cty}
\end{figure}

>>>>

  \begin{center}
\vbox{\hbox{\hspace{1cm}
    \includegraphics[width=0.8\textwidth]{diagrams/compar-dg.eps}}
   \hbox{ \hspace{1cm} \includegraphics[width=0.8\textwidth]{diagrams/compar-dg-bdt.eps}}}
    \caption{Advection problem test-cases: control volume solution (5
      finite elements with constant width of 0.2 are used) with
      Courant number of 0.005 (top) and 0.5 (bottom).  The solution
      has been advected a distance of 0.2 with 200 (top) and 2
      (bottom) time steps. The 50 elements solution is shown to give
      an indication of the accuracy of the solutions and has a Courant
      number (based on the element width) of 0.005.\label{compar-dg}}
  \end{center}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=1.2\textwidth]{diagrams/compar-dg-bdt-diff}
\end{center}
\caption{Advection problem test-cases: discontinuous (between
  elements) control volume formulation within 5 finite elements (width
  of 0.2). Courant number is of 0.5 (or 2 if based on the minimum
  width of a control volume). The solution has been advected a
  distance of 0.2 with 2 time steps. The 50 element solution is for
  pure advection and is shown to give an indication of the accuracy of
  the solutions and has a Courant number (based on the element width)
  of 0.005.\label{compar-dg-bdt-diff}}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\vbox{\hbox{\hspace{-1.3cm}
\includegraphics[width=17.5cm,height=12.5cm]{diagrams/theta-bdt}}
\vspace{0.cm}}
\caption{Advection problem test-cases: value of $\theta$-time stepping
  parameter on the faces of the control volumes at the end of the
  simulation -- 5 FE are used and the Courant number (based on the
  element width of 0.2) is 0.5 (or 2 if based on the minimum width of
  a control volume). The solution has been advected a distance of 0.2
  length units with 2 time steps.}
\label{theta-bdt}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\vbox{\vspace{-2cm}
\hbox{\hspace{-.5cm}
\includegraphics[width=14.0cm,height=10.cm]{diagrams/converg-cv}}
\vspace{-.5cm}
\hbox{\hspace{-.5cm}
\includegraphics[width=14.0cm,height=10.cm]{diagrams/converg-fem}}
}\vspace{-.3cm}
\caption{Advection problem test-cases: convergence of the solutions
  with discontinuity between elements and with increased resolution. A
  pure advection problem (from left to right) - the Courant number is
  0.005. The solution has been advected a distance of 0.2. Control
  volume solutions (top) and FEM interpolation (bottom) of the control
  volume solutions are shown. \label{converg}}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\hbox{\hspace{-1cm}
\includegraphics[width=17.5cm,height=12.5cm]{diagrams/converg-compare-fem}}
\caption{Advection problem test-cases: comparison of convergence of
  the solutions with and without discontinuity between elements and
  with an upwind and central difference flux between the elements and
  with increased resolution (dashed line: 5 elements, full line: 20
  elements). A pure advection problem (from left to right) - the
  Courant number is 0.005. The solution has been advected a distance
  of 0.2. The FEM interpolation of the control volume solutions are
  shown. \label{converg-compare-fem}}
\end{figure}

@@

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=1.\textwidth]{diagrams/bl-exact-meth-upwind.eps}
\end{center}
\caption{Buckley--Leverett test-cases: Saturation solutions for the
  continuous upwind method for different 1D P$_{1}$DG-P$_{2}$ mesh
  resolutions and comparison against standard analytical solution.
\label{bl-exact-meth-upwind}}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
  %\begin{center}
\vbox{\hbox{\hspace{2.5cm}
    \includegraphics[width=0.62\textwidth]{diagrams/BL_1d_P0DGP1_convergence.eps}}
\vspace{-.0cm}\hbox{\hspace{2.5cm}
    \includegraphics[width=0.62\textwidth]{diagrams/BL_1d_P1DGP2_convergence.eps}}
\vspace{-.0cm}\hbox{\hspace{2.5cm}
    \includegraphics[width=0.62\textwidth]{diagrams/BL_1d_P2DGP3_convergence.eps}}}
   % \includegraphics[width=0.45\textwidth]{BL_2d_P1DGP2_convergence}
    \caption{Buckley--Leverett test-cases: Saturation profiles for a
      number of element pairs and numerical resolutions in 1D --
      P$_{0}$DG-P$_{1}$ (top), P$_{1}$DG-P$_{2}$ and P$_{2}$DG-P$_{3}$
      (bottom).\label{fig:BL_profiles}}
  %\end{center}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\vbox{\hbox{\hspace{1.cm}
    \includegraphics[width=0.8\textwidth]{diagrams/L1_convergence_rate.eps}}
\vspace{.0cm}\hbox{\hspace{1.cm}
    \includegraphics[width=0.8\textwidth]{diagrams/L2_convergence_rate.eps}}}
    \caption{Buckley--Leverett test-cases: L1 (top) and L2 (bottom)
      error convergence rates for a number of element
      pairs. \label{fig:BL_converg-rates}}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=1.\textwidth]{diagrams/bl-upwind-v-up-and-down.eps}
\end{center}
\caption{Buckley--Leverett test-cases: Comparison of the optimal
  upwind formulation when using upwinding (OU) and coupled
  upwind/downwind (OU-D). The finite element interpolation of the
  saturation field $\left(S_{1}\right)$ is shown at different mesh
  resolutions. Downwind seems to detract from the accuracy of the
  solution. \label{bl-upwind-v-up-and-down}}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\vbox{
\begin{center}
\includegraphics[width=1.\textwidth]{diagrams/bl-exact-meth-cv-0-8-ele50.eps}
\end{center}
\vspace{0.cm}}
\caption{Buckley--Leverett test-cases: Comparison of control volume
  solutions using 80$\%$ upwinding and with optimal upwinding and
  using 50 continuous P$_{1}$DG-P$_{2}$
  elements. \label{bl-exact-meth-cv-0-8-ele50}}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=1.\textwidth]{diagrams/bl-dg-2eles.eps}
\end{center}
\caption{Buckley--Leverett test-cases: Two element solution using the
  discontinuous formulation. Saturation field from both CV solution
  and FEM interpolation are shown.  \label{bl-dg-2eles}}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\vbox{
\hbox{\hspace{.3cm}\includegraphics[width=.9\textwidth]{diagrams/bl-dg-4-10-20.eps}}
\vspace{-0.cm}
\hbox{\hspace{.3cm}\includegraphics[width=.9\textwidth]{diagrams/bl-dg-cent-4-10-20.eps}}}
\caption{Buckley--Leverett test-cases: Saturation field obtained from
  the discontinuous and continuous formulation with different mesh
  resolutions. Solutions with (top) and without (bottom) upwinding
  scheme. Notice that oscillations are suppressed with the upwinding
  scheme.\label{bl-dg-cent-4-10-20}}
\end{figure}

%%%
%%%  FIGURE 
%%%
\begin{figure}[h]
\vbox{
\hbox{\hspace{.3cm}\includegraphics[width=.9\textwidth]{diagrams/bl-dg-4-10-vers-cty.eps}}
\vspace{-0.cm}
\hbox{\hspace{.3cm}\includegraphics[width=.9\textwidth]{diagrams/bl-dg-p1-2-4-5-10-20-40.eps}}}
\caption{Buckley--Leverett test-cases: Saturation field obtained from
  (top) continuous and discontinuous (between elements) formulations
  (solution with 50 elements may be considered as a converged
  result). Solution obtained (bottom) from linear pressure (P1)
  formulation with different mesh resolution with comparison against
  P2-pressure formulation (continuous). \label{bl-dg-4-10-vers-cty}}
\end{figure}

\begin{figure}[H]
\vbox{
\begin{center}
\includegraphics[width=17.5cm,height=12.5cm]{diagrams/bl-dg-4-10-vers-cty}
\end{center}
\vspace{0.cm}}
\caption{Gas saturations shown comparing the accuracy of the
  discontinuous between elements and continuous formulation. The 50
  element continuous solution may be viewed as a converged result.  }
\label{bl-dg-4-10-vers-cty}
\end{figure}
